% file: appendix_b.tex
% 	CORE Library
% 	$Id: appendix_c.tex,v 1.1.1.1 2007/04/09 23:35:54 exact Exp $ 

\newpage
\section{APPENDIX: Brief History}

\begin{description}
\item{Version 1.1}
(Dec 1998) The initial implementation
by Karamcheti, Li, Pechtchanski and Yap \cite{klpy:core:98}
was based on the \rexpr\ package,
designed by Dub\'e and Yap (circa 1993)
and rewritten by Ouchi \cite{ouchi:thesis}.  
Details about the underlying 
algorithms (especially in \BF) and their error analysis
may be found in the Ouchi's thesis \cite{ouchi:thesis}.

\item{Version 1.2}
(Sep 1999) Incorporates the BFMS root bound and other 
techniques to give significant speedup to the system. 

\item{Version 1.3}
(Sep 2000) Two improvements (new root bounds
and faster big number packages based on \lidia/\cln)
gave significant general speedup.  
More examples, including a hypergeometric function package
and a randomized geometric theorem prover.

\item{Version 1.4}
(Sep 2001) 
Introduced a floating point filter based on the BFS filter,
incremental square root computation,
improved precision-sensitive evaluation algorithms,
better numerical I/O support.
Our big integer and big rational packages
are now based on \gmp, away from \lidia/\cln.

\item{Version 1.5}
(Aug 2002) Improvements in speed from better root bounds (k-ary bounds),
CGAL compatibility changes, file I/O for large mathematical constants
(BigInt, BigFloat, BigRat), improved hypergeometric package.

\item{Version 1.6}
(June 2003)  The introduction of real algebraic numbers (a first
among such systems).
CORE is now issued under the the Q PUBLIC LICENSE (QPL),
concurrent with its being distributed with CGAL under commercial
licenses by Geometry Factory, the CGAL commercial spin-off.
Incorporated Polynomial and Sturm classes into Core Library.

\item{Version 1.7}
(Aug 2004) Introduced algebraic curves and bivariate polynomials.
An interactive version of Core Library called "InCore" is available.
Beginning basic graphic capability for display of curves.  
Restructuring of number classes (Expr, BigFloat, etc)
to have common reference counting and rep facilities.

\end{description}


